---
title: "Project_2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(cowplot)
```

## Importing the data

```{r}
countries <- read.csv("countries of the world.csv")
dim(countries)
str(countries)
head(countries)
```

## Cleaning data

```{r}
# change decimals from commas to periods and treat them as numbers instead of factors
countries[3:20] <- apply(apply(countries[3:20], 2, gsub, patt=",", replace="."), 2, as.numeric)
# trim whitespace from text values in dataframe (region names had right-side filler)
countries[1:2] <- apply(countries[1:2], 2, trimws)
# check how many countries are missing data
which(!complete.cases(countries))
mean(complete.cases(countries))
```

## Basic Data Analysis

#### Number summaries

```{r}
summary(countries)
count(countries, Region)
```

#### Sorting countries and regions by total population

```{r}
countries[c(1, 3)] %>%
  mutate(Percentage = 100 * Population / sum(Population)) %>%
  arrange(desc(Population)) %>%
  head(10)

countries %>%
  group_by(Region) %>%
  summarize(Total_Population = sum(Population)) %>%
  mutate(Percentage = 100 * Total_Population / sum(Total_Population)) %>%
  arrange(desc(Total_Population))
```

#### Sorting countries and regions by total area

```{r}
countries[c(1, 4)] %>%
  mutate(Percentage = 100 * Area..sq..mi.. / sum(Area..sq..mi..)) %>%
  arrange(desc(Area..sq..mi..)) %>%
  head(10)

countries %>%
  group_by(Region) %>%
  summarize(Total_Area = sum(Area..sq..mi..)) %>%
  mutate(Percentage = 100 * Total_Area / sum(Total_Area)) %>%
  arrange(desc(Total_Area))
```

## Data Visualization

#### Plotting countries by population and GDP per capita, grouped by region

```{r}
countries %>%
  ggplot(mapping = aes(x = log10(GDP....per.capita.), y = log10(Population), color = factor(Region))) +
  geom_smooth(method = 'lm', alpha = 0.10) +
  geom_point(alpha = 0.60)
```

#### Plotting countries by GDP per capita and literacy, grouped by region

```{r}
countries %>% subset(!is.na(Literacy....), !is.na(GDP....per.capita.)) %>%
  ggplot(mapping = aes(x = Literacy...., y = log10(GDP....per.capita.), color = factor(Region))) +
  geom_smooth(method = 'lm', alpha = 0.10) +
  geom_point(alpha = 0.60)
```

#### Plotting countries by proportion of GDP from each GDP sector against GDP per capita

```{r}
plot1 <- ggplot(subset(countries, !is.na(Agriculture)), aes(x = Agriculture, y = log10(GDP....per.capita.))) +
  geom_vline(xintercept = mean(countries$Agriculture, na.rm = TRUE)) +
  geom_smooth(method = 'lm', alpha = 0.5) +
  geom_point(alpha = 0.3) +
  ylim(2.6, 4.8) +
  xlim(0, 1)

plot2 <- ggplot(subset(countries, !is.na(Industry)), aes(x = Industry, y = log10(GDP....per.capita.))) +
  geom_vline(xintercept = mean(countries$Industry, na.rm = TRUE)) +
  geom_smooth(method = 'lm', alpha = 0.5) +
  geom_point(alpha = 0.3) +
  ylim(2.6, 4.8) +
  xlim(0, 1)

plot3 <- ggplot(subset(countries, !is.na(Service)), aes(x = Service, y = log10(GDP....per.capita.))) +
  geom_vline(xintercept = mean(countries$Service, na.rm = TRUE)) +
  geom_smooth(method = 'lm', alpha = 0.5) +
  geom_point(alpha = 0.3) +
  ylim(2.6, 4.8) +
  xlim(0, 1)

plot_grid(plot1, plot2, plot3, ncol = 3)
```

#### Plotting regions by total area and total population

```{r}
set.seed(1) # force same text repel every time
countries %>% group_by(Region) %>%
  summarize(Total_Area = sum(Area..sq..mi.., na.rm = TRUE), Total_Population = sum(Population, na.rm = TRUE)) %>%
  ggplot(mapping = aes(x = log10(Total_Area), y = log10(Total_Population), label = Region)) +
  geom_smooth(method = 'lm') +
  geom_point(color = 'red') +
  geom_text_repel()
```

#### Plotting regions by mean GDP per capita and mean literacy

```{r}
set.seed(1) # force same label repel every time
countries %>% group_by(Region) %>%
  summarize(Mean_Literacy = mean(Literacy...., na.rm = TRUE), Mean_GDP = mean(GDP....per.capita., na.rm = TRUE)) %>%
  ggplot(mapping = aes(x = Mean_Literacy, y = Mean_GDP, label = Region)) +
  geom_smooth(method = 'lm') +
  geom_point() + 
  geom_label_repel(
    arrow = arrow(length = unit(0.02, "npc"), type = "closed", ends = "first"),
    force = 10
  )
```

#### Boxplot of mean infant mortality rate per 1000 births, grouped by region

```{r}
countries %>% group_by(Region) %>% subset(!is.na(Infant.mortality..per.1000.births.)) %>%
  ggplot(mapping = aes(x = Region, y = Infant.mortality..per.1000.births., fill = Region)) +
  geom_boxplot(alpha=0.6) +
  theme(legend.position = "none") +
  coord_flip()
```